WSClean: an implementation of a fast, generic wide-field imager

### 2014-02-28 articles/2014/Article-WSClean/wsclean.tex
Replace noise measurement with better one, final touches to article before submit to collaboration

\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{\hspace{2cm}E-mail: \url{andre.offringa@anu.edu.au}},
B.~McKinley\ANU$^,$\CAASTRO,
N.~Hurley-Walker\Curtin,
F.~H.~Briggs$^{1,2}$, \newauthor
R.~B.~Wayth\Curtin$^,$\CAASTRO, 
D.~L.~Kaplan\UWisc,
M.~E.~Bell\USydney$^,$\CAASTRO,
L.~Feng\MIT,
A.~R.~Neben\MIT,
J.~D.~Hughes\ANU, \newauthor
J.~Rhee\ANU,
G.~Bernardi\SKASA,
J.~D.~Bowman\ASU,
R.~J.~Cappallo\Haystack,
B.~E.~Corey\Haystack, \newauthor
A.~A.~Deshpande\RRI, 
D.~Emrich\Curtin,
B.~M.~Gaensler\USydney$^,$\CAASTRO,
R.~Goeke\MIT,
L.~J.~Greenhill\CfA, \newauthor
B.~J.~Hazelton\UW, 
J.~N.~Hewitt\MIT,
M.~Johnston-Hollitt\Victoria,
J.~C.~Kasper\CfA, \newauthor
E.~Kratzenberg\Haystack, 
C.~J.~Lonsdale\Haystack,
M.~J.~Lynch\Curtin,
S.~R.~McWhirter\Haystack,
D.~A.~Mitchell\CASS$^,$\CAASTRO, \newauthor
M.~F.~Morales\UW, 
E.~Morgan\MIT,
D.~Oberoi\Tata,
S.~M.~Ord\Curtin$^,$\CAASTRO,
T.~Prabu\RRI, 
A.~E.~E.~Rogers\Haystack, \newauthor
D.~A.~Roshi\NRAO \footnote{The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc.},
N.~Udaya~Shankar\RRI,
K.~S.~Srivani\RRI,
R.~Subrahmanyan\RRI$^,$\CAASTRO, \newauthor
S.~J.~Tingay\Curtin$^,$\CAASTRO, 
M.~Waterson\Curtin$^,$\ANU,
R.~L.~Webster\UMelbourne$^,$\CAASTRO,
A.~R.~Whitney\Haystack, \newauthor
A.~Williams\Curtin,
C.~L.~Williams\MIT
\\
\ANU{}Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics (CAASTRO), Australian National University, Canberra, ACT 2611, Australia \\
\Curtin{}International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
\UWisc{}Department of Physics, University of Wisconsin--Milwaukee, Milwaukee, WI 53201, USA\\
\MIT{}Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
\SKASA{}Square Kilometre Array South Africa (SKA SA), Cape Town 7405, South Africa\\
\ASU{}School of Earth and Space Exploration, Arizona State University, Tempe, AZ 85287, USA\\
\Haystack{}MIT Haystack Observatory, Westford, MA 01886, USA\\
\RRI{}Raman Research Institute, Bangalore 560080, India\\
\USydney{}Sydney Institute for Astronomy, School of Physics, The University of Sydney, NSW 2006, Australia\\
\CfA{}Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
\UW{}Department of Physics, University of Washington, Seattle, WA 98195, USA\\
\Victoria{}School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
\CASS{}CSIRO Astronomy and Space Science, Marsfield, NSW 2122, Australia\\
\Tata{}National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India\\
\NRAO{}National Radio Astronomy Observatory, Charlottesville and Greenbank, USA\\
\UMelbourne{}School of Physics, The University of Melbourne, Parkville, VIC 3010, Australia\\
}

@@

\begin{abstract}
We present a new interferometric imager that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm. The performance dependencies of CASA's $w$-projection and our new imager are analysed and analytical functions are derived that describe the required computing cost for both imagers. On MWA data, we find our new method to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution and with full polarization correction.  We predict the computing costs for several other arrays and estimate that our imager is a factor of 2--12 faster, depending on the array configuration. Imaging SKA observations is challenging, as we estimate the computing cost of full SKA1 low imaging to be 60 PetaFLOPS with current techniques. We find that combining $w$-stacking with the $w$-snapshot algorithm does not significantly improve computing requirements. The source code of our new imager is publicly released.
\end{abstract}

@@

There are several methods to deal with the $w$-term during imaging: facetting \citep{facetting-cornwell}, a three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}, $w$-projection \citep{wprojection-cornwell}, $w$-stacking \citep{ska-memo-regridding-2011} and $w$-snapshots \citep{widefield-imaging-ska-cornwell}.

@@

Another option exists for imaging MWA data: the Real-Time System (RTS, \citealt{rts-mwa}). This has been designed as an efficient calibration and imaging pipeline specifically for MWA data. It can use GPUs to improve efficiency, and uses a $w$-projection kernel during imaging to correct for the $w$-term. The kernel can be kept small, because the data is processed in snapshots that have been phase-centered to zenith. The RTS can not perform deconvolution as of yet.

@@

In this article, we present a new implementation of a generic wide-field imager that is significantly faster than CASA's $w$-projection implementation, and is able to succesfully image any MWA observation. To obtain the increase in speed, the implementation uses the $w$-stacking method for correcting the $w$-terms, optionally combined with $w$-snapshot imaging. We named the new imager ``WSClean'', as an abbreviation for ``$w$-Stacking Clean''. Our new imaging implementation, which we believe to be the fastest generic widefield imager available, is publicly released\footnote{TODO add URL}.

@@

An interferometer samples the complex visibility function
\begin{align}\notag
V(u,v,w) = & \iint \frac{A(l,m) I(l,m)}{\sqrt{1-l^2-m^2}} \times \\ \label{eq:visibility-function}
& e^{-2\pi i \left(ul + vm + w(\sqrt{1-l^2-m^2}-1)\right)} dl dm,
\end{align}
where $u,v,w$ is a baseline coordinate in the coordinate system of the array, $A$ is the primary-beam function, $I$ is the sky function and $l,m$ are cosine sky coordinates. We will use $I'(l,m)$ to denote the sky function before primary-beam correction, $I'(l,m)=A(l,m)I(l,m)$. We will not discuss calibration, but assume $V$ has been calibrated before imaging. In the case of a polarized measurement, the symbols become $2\times 2$ matrices, but without loss of generality we will ignore polarization and treat inversion as a scalar problem. Imaging consists of inverting Eq.~\eqref{eq:visibility-function}, i.e., to find $I'$ from $V$.

@@

To derive the $w$-stacking technique, Eq.~\eqref{eq:visibility-function} is rewritten to
\begin{align}\notag
V(u,v,w) = & \iint \frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} \times \\ \notag
& e^{-2\pi i \left(ul + vm\right)} dl dm.
\end{align}
This is an ordinary two-dimensional Fourier transform going from $u,v$ space to $l,m$ space, and can be inverted to get:
% Intermediate step:
%\begin{align}\notag
%\frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} = & \iint V(u,v,w) \times \\ \notag
%& e^{-2\pi i \left(ul + vm\right)} du dv,
%\end{align}
%which can be rewritten to
\begin{align}\notag
\frac{I'(l,m)}{\sqrt{1-l^2-m^2}} = & e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \iint V(u,v,w) \times \\ \notag
& e^{2\pi i \left(ul + vm\right)} du dv.
\end{align}
Integrating both sides over $w_{\min}$ to $w_{\max}$, the minimum and maximum value of $w$, results in
\begin{align}\notag
\frac{I'(l,m)\left(w_{\max} - w_{\min}\right)}{\sqrt{1-l^2-m^2}} = \int\limits_{w_{\min}}^{w_{\max}} e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \times \\ \label{eq:wstacking}
\iint V(u,v,w)  e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
The final step is to make the $u,v,w$ parameters discrete, so that the integration over $u$ and $v$ can become an inverse FFT and the integration over $w$ becomes a summation. This shows that the sky function can be reconstructed by i)~grid samples with equal $w$-value on a uniform grid; ii)~calculating the inverse FFT; iii)~apply the direction-dependent phaseshift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; iv)~repeat this for all $w$-values and add the results together; v)~apply the final scaling.

@@

The inverse of imaging, i.e., to calculate the visibility from a model image, can be done by reversing the $w$-stacking algorithm: i)~inverse scale the image; ii)~copy the image to several layers; iii)~inverse apply the direction-dependent phaseshift for each layer; iv)~FFT each layer; and v)~sample a required visibility from the correct $w$-layer. We will refer to this operation as prediction.

@@

\subsection{Time complexity of $w$-stacking} \label{sec:time-complexity}
It is useful to analyse the time complexity of $w$-stacking and compare it with $w$-projection, to understand which algorithm performs better in a given situation. We will use the following symbols: $N_\textrm{wlay}$ is the number of $w$-layers for $w$-stacking, $N_\textrm{pix}$ is the number of pixels in the image along each dimension, $N_\textrm{vis}$ is the number of visibilities, $N_\textrm{kern}$ is the size of the antialiasing kernel (see \S\ref{sec:gridding}), $N_\textrm{wkern}$ is the size of the $w$-kernel for $w$-projection, $w_{\max}$ is the maximum $w$ value and $\alpha_\textrm{FOV}$ is the imaging field of view.

@@

\section{The WSClean imager implementation} \label{sec:implementation}
With the purpose of testing new algorithms for the imaging of MWA data, we have written a new imager around the $w$-stacking algorithm. As will be shown in Sect.~\ref{sec:application-to-nonmwa}, the imager is not specialized for the MWA, and has been succesfully used for imaging data from other telescopes.

The new imager, called ``WSClean'', is written in the C++ language. It reads visibilities from CASA measurement sets and writes output images to Flexible Image Transport System (FITS) files. Several steps are multi-threaded using the threading module of the C++11 standard library. These are: reading and writing; gridding from and to different $w$-layers; performing the FFTs; and H\"ogbom Clean iterations (peak-finding and image subtraction, \citealt{hogbom-clean}). For the latter, intrinsics are used as well. Because of these optimizations, minor Cleaning iterations with an image size of $3072\times3072$ are performed at a rate of hundreds per second. This is fast enough to make the Clark Clean optimization \citep{clark-clean}, which consists of considering only a subset of pixels with a reduced PSF, less relevant. However, the PSF varies for imaging with non-zero $w$-values, and subtracting a constant PSF in image space leads to inaccuracies. After a number of minor iterations, it is therefore beneficial to invert the model back to the visibilities via prediction, and subtract the model directly from the visibilities \citep{wprojection-cornwell}. This is similar to the Cotton-Schwab cleaning method \citep{cotton-schwab-clean}. WSClean allows cleaning individual polarizations, or can jointly deconvolve the polarizations. In the latter case, peaks are searched in $pp^2+2pq \overline{(pq)} + qq^2$ space, where $\overline{pq}=qp$ is the complex conjugate of $pq$, and the PSF is subtracted from the polarizations with different factors.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=13.5cm]{img/fullsky-together}
\caption{A beam-corrected MWA image of a 112~s zenith observation that covers almost the full sky. The lower image shows a zoom-in on the southern side lobe. PKS J2358-6054 ($\sim100$ Jy) is visible in the centre of the southern side lobe and resolved. The noise level in the side lobe is about 200 mJy.}
\label{fig:full-sky-example}
\end{center}
\end{figure*}

@@

To efficiently image the full sky, an observation is split into short snapshots, their phase centre is changed to zenith and subsequently a suitable area is imaged. Depending on desired resolution and dynamic range, this might require very large images. To make an image at the MWA resolution, the image needs to have approximately 10k pixels in width and height. The $w$-values will be very small at zenith, because at zenith only the vertical offsets of the antennas will contribute to the $w$-value. For the MWA, the maximum height difference between tiles is $8.5$m. The tiles are in fact on a slight slope, and by fitting a plane to the antennas and changing the phase centre towards the normal of that plane, the maximum $w$-value decreases to $5.5\textrm{m} / \lambda$. In the following sections, when discussing the MWA zenith direction we are referring to this optimal $w$-direction.

@@

An example of the difference in projection between non-recentred and recentred imaging is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13~min MWA observation imaged with WSClean. The processing steps performed for this image were i) preprocessing of the observation with the Cotter preprocessing pipeline, which include time averaging and RFI flagging with the AOFlagger \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}; ii) calibrate on Hydra~A without direction dependence using a custom implementation of the RTS full-polarization calibration algorithm \citep{rts-mwa}; and iii) image each of the seven 112 s intervals separately. Cleaning an image that is in a different projection yields slightly different results, but qualitativily the two images are clearly of equal accuracy. The wall-clock time for imaging is 60 min and 41 min for respectively normal projection and the recentred image. Of the 41 min, 3 min is spent on phase rotating the visibilities.

Another method to shift an image is by using the periodicity of the FFT function. Since sources outside the field of view will be aliased back into the field, the image can be transformed to have the alias in the centre. This complicates gridding during imaging, because the antialiasing kernel needs to have a pass-band shape instead of a low-pass shape. Therefore, we chose to implement the former method.

\subsection{Beam correction}
\begin{figure*}
\begin{center}
\includegraphics[width=6cm]{img/pulsar-stokesi}\hspace{1cm}\includegraphics[width=6cm]{img/pulsar-stokesv}
\caption{Stokes I (left panel, total power) and Stokes V (right panel, circular polarization) images of PSR J0437-4715, demonstrating the full-polarization beam correction capability for MWA observations with WSClean. The pulsar (centre of image) displays 17\% circular polarization.}
\label{fig:pulsar-stokes-iv}
\end{center}
\end{figure*}
Because the MWA consists of fixed, beamformed dipole antennas, the voltage beam of a MWA tile is described by a complex, non-diagonal Jones matrix. Alternatively, Mueller matrices can be used as well, but in general these result in more computations \citep{revisiting-me-i}. When assuming all individual antennas of the tiles are working properly, all tiles have the same beam. The beam can then be corrected in image space for all tiles at once with
\begin{equation}\label{eq:beam-correction}
 I(l, m) = J(l,m)^{-1} I'(l, m) \left(J(l,m)^*\right)^{-1},
\end{equation}
where each term is a $2\times2$ complex matrices, with $J$ the voltage beam and $^*$ denoting the conjugate transpose. Because $J$ is complex and non-diagonal, all 4 combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such. However, this is incorrect because they are not necessarily orthogonal, i.e., ($2\times$) Stokes I$\neq pp + qq$.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters. The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry. To make proper MWA beam correction possible, WSClean can output both real $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images. This requires four runs of the algorithm. Once these images are created, a separate program is used to perform the correction of Eq.~\eqref{eq:beam-correction}. The method is demonstrated in Fig.~\ref{fig:pulsar-stokes-iv}, which displays a detection of pulsar J0437-4715 in Stokes V. The image was made with an initial pipeline for the radio-sky monitor project that uses the MWA to search for transients and exoplanets. The pipeline includes WSClean to make the Stokes-parameter images.

@@

\subsection{Decreasing inversion resolution} \label{sec:decreasing-inversion-resolution}
Typically, for cleaning it is desired that images have a pixel size at least 5 times smaller than the synthesized-beam width. This improves the accuracy of the clean algorithm, because the positions of image maxima will be closer to the actual source positions. This factor of $\sim$5 is normally taken into account in the overal imaging resolution, i.e., the image size is increased during inversion. For the $w$-projection method, the cost of inversion with an increased resolution is small, because it does not increase the size of the $w$-kernels. It will affect the inverse FFT, but the relative cost of this step is negligable in the total cost. The $w$-stacking method is however significant affected by the image resolution, because it performs many inverse FFTs.

The spatial frequencies in the output image are band limited by the synthesized beam. Therefore, it is not required to perform the inversion at higher resolution. As long as the $uv$-plane is sampled with at least the Nyquist frequency, a high resolution image can be perfectly reconstructed from an inversion at lower resolution. After an upscaled image has been cleaned, a model is created at the same (high) resolution. Because the model consists of delta functions, the spatial frequencies in the model are not band limited. Therefore, downscaling the model will remove information from the model. However, only the low spatial frequencies of this image will be used in the prediction. As long as downscaling does not modify the low-frequency components, the output of the prediction step will not change. This requires low-pass filtering the model before decimation.

>>>>

\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{E-mail: \url{andre.offringa@anu.edu.au}},
B.~McKinley\ANU$^,$\CAASTRO,
N.~Hurley-Walker\Curtin,
F.~H.~Briggs$^{1,2}$, \newauthor
R.~B.~Wayth\Curtin$^,$\CAASTRO, 
D.~L.~Kaplan\UWisc,
M.~E.~Bell\USydney$^,$\CAASTRO,
L.~Feng\MIT,
A.~R.~Neben\MIT,
J.~D.~Hughes\ANU, \newauthor
J.~Rhee\ANU,
G.~Bernardi\SKASA,
J.~D.~Bowman\ASU,
R.~J.~Cappallo\Haystack,
B.~E.~Corey\Haystack, \newauthor
A.~A.~Deshpande\RRI, 
D.~Emrich\Curtin,
B.~M.~Gaensler\USydney$^,$\CAASTRO,
R.~Goeke\MIT,
L.~J.~Greenhill\CfA, \newauthor
B.~J.~Hazelton\UW, 
J.~N.~Hewitt\MIT,
M.~Johnston-Hollitt\Victoria,
J.~C.~Kasper\CfA, \newauthor
E.~Kratzenberg\Haystack, 
C.~J.~Lonsdale\Haystack,
M.~J.~Lynch\Curtin,
S.~R.~McWhirter\Haystack,
D.~A.~Mitchell\CASS$^,$\CAASTRO, \newauthor
M.~F.~Morales\UW, 
E.~Morgan\MIT,
D.~Oberoi\Tata,
S.~M.~Ord\Curtin$^,$\CAASTRO,
T.~Prabu\RRI, 
A.~E.~E.~Rogers\Haystack, \newauthor
D.~A.~Roshi\NRAO, %\footnote{The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc.},
N.~Udaya~Shankar\RRI,
K.~S.~Srivani\RRI,
R.~Subrahmanyan\RRI$^,$\CAASTRO, \newauthor
S.~J.~Tingay\Curtin$^,$\CAASTRO, 
M.~Waterson\Curtin$^,$\ANU,
R.~L.~Webster\UMelbourne$^,$\CAASTRO,
A.~R.~Whitney\Haystack, \newauthor
A.~Williams\Curtin,
C.~L.~Williams\MIT
\\
\ANU{}Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics (CAASTRO), Australian National University, Canberra, ACT 2611, Australia \\
\Curtin{}International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
\UWisc{}Department of Physics, University of Wisconsin--Milwaukee, Milwaukee, WI 53201, USA\\
\MIT{}Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
\SKASA{}Square Kilometre Array South Africa (SKA SA), Cape Town 7405, South Africa\\
\ASU{}School of Earth and Space Exploration, Arizona State University, Tempe, AZ 85287, USA\\
\Haystack{}MIT Haystack Observatory, Westford, MA 01886, USA\\
\RRI{}Raman Research Institute, Bangalore 560080, India\\
\USydney{}Sydney Institute for Astronomy, School of Physics, The University of Sydney, NSW 2006, Australia\\
\CfA{}Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
\UW{}Department of Physics, University of Washington, Seattle, WA 98195, USA\\
\Victoria{}School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
\CASS{}CSIRO Astronomy and Space Science, Marsfield, NSW 2122, Australia\\
\Tata{}National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India\\
\NRAO{}National Radio Astronomy Observatory, Charlottesville and Greenbank, USA\\
\UMelbourne{}School of Physics, The University of Melbourne, Parkville, VIC 3010, Australia\\
}

@@

\begin{abstract}
We present a new interferometric imager that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm. The performance dependencies of CASA's $w$-projection and our new imager are analysed and analytical functions are derived that describe the required computing cost for both imagers. On MWA data, we find our new method to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution and with full polarization correction.  We predict the computing costs for several other arrays and estimate that our imager is a factor of 2--12 faster, depending on the array configuration. We estimate the computing cost for imaging full SKA1-low observations to be 60 PetaFLOPS with current techniques. We find that combining $w$-stacking with the $w$-snapshot algorithm does not significantly improve computing requirements. The source code of our new imager is publicly released.
\end{abstract}

@@

There are several methods to deal with the $w$-term during imaging: faceting \citep{facetting-cornwell}, a three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}, $w$-projection \citep{wprojection-cornwell}, $w$-stacking \citep{ska-memo-regridding-2011} and $w$-snapshots \citep{widefield-imaging-ska-cornwell}.

@@

Another option exists for imaging MWA data: the Real-Time System (RTS, \citealt{rts-mwa}). This has been designed as an efficient calibration and imaging pipeline specifically for MWA data. It can use GPUs to improve efficiency, and uses a $w$-projection kernel during imaging to correct for the $w$-term. The kernel can be kept small, because the data is processed in snapshots. The RTS can not perform deconvolution as of yet.

@@

In this article, we present a new implementation of a generic wide-field imager that is significantly faster than CASA's $w$-projection implementation, and is able to successfully image any MWA observation. To obtain the increase in speed, the implementation uses the $w$-stacking method for correcting the $w$-terms, optionally combined with a new technique for $w$-snapshot imaging. We named the new imager ``WSClean'', as an abbreviation for ``$w$-Stacking Clean''. Our new imaging implementation, which we believe to be the fastest generic widefield imager available, is publicly released\footnote{The WSClean source code can be found at:\\\url{http://sourceforge.net/p/wsclean}}.

@@

An interferometer samples the complex visibility function
\begin{align}\notag
V(u,v,w) = & \iint \frac{A(l,m) I(l,m)}{\sqrt{1-l^2-m^2}} \times \\ \label{eq:visibility-function}
& e^{-2\pi i \left(ul + vm + w(\sqrt{1-l^2-m^2}-1)\right)} dl dm,
\end{align}
where $u,v,w$ is a baseline coordinate in the coordinate system of the array, $A$ is the primary-beam function, $I$ is the sky function and $l,m$ are cosine sky coordinates. We will use $I'(l,m)$ to denote the sky function before primary-beam correction, $I'(l,m)=A(l,m)I(l,m)$. We will not discuss calibration, but assume $V$ has been calibrated before imaging. In the case of a polarized measurement, the symbols become $2\times 2$ matrices and beam correction is more complicated, but without loss of generality we will ignore polarization and treat inversion as a scalar problem. Imaging consists of inverting Eq.~\eqref{eq:visibility-function}, i.e., to find $I'$ from $V$.

@@

To derive the $w$-stacking technique, Eq.~\eqref{eq:visibility-function} is rewritten to
\begin{align}\notag
V(u,v,w) = & \iint \frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} \times \\ \notag
& e^{-2\pi i \left(ul + vm\right)} dl dm.
\end{align}
This is an ordinary two-dimensional Fourier transform going from $u,v$ space to $l,m$ space, and can be inverted to get:
% Intermediate step:
%\begin{align}\notag
%\frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} = & \iint V(u,v,w) \times \\ \notag
%& e^{-2\pi i \left(ul + vm\right)} du dv,
%\end{align}
%which can be rewritten to
\begin{align}\notag
\frac{I'(l,m)}{\sqrt{1-l^2-m^2}} = & e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \iint V(u,v,w) \times \\ \notag
& e^{2\pi i \left(ul + vm\right)} du dv.
\end{align}
Integrating both sides over $w_{\min}$ to $w_{\max}$, the minimum and maximum value of $w$, results in
\begin{align}\notag
\frac{I'(l,m)\left(w_{\max} - w_{\min}\right)}{\sqrt{1-l^2-m^2}} = \int\limits_{w_{\min}}^{w_{\max}} e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \times \\ \label{eq:wstacking}
\iint V(u,v,w)  e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
The final step is to make the $u,v,w$ parameters discrete, so that the integration over $u$ and $v$ can become an inverse FFT and the integration over $w$ becomes a summation. This shows that the sky function can be reconstructed by i)~gridding samples with equal $w$-value on a uniform grid; ii)~calculating the inverse FFT; iii)~apply the direction-dependent phase shift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; iv)~repeat this for all $w$-values and add the results together; v)~apply the final scaling.

@@

The inverse of imaging, i.e., to calculate the visibility from a model image, can be done by reversing the $w$-stacking algorithm: i)~inverse scale the image; ii)~copy the image to several layers; iii)~inverse apply the direction-dependent phase shift for each layer; iv)~FFT each layer; and v)~sample a required visibility from the correct $w$-layer. We will refer to this operation as prediction.

@@

\subsection{Time complexity of $w$-stacking} \label{sec:time-complexity}
It is useful to analyse the time complexity of $w$-stacking and compare it with $w$-projection, to understand which algorithm performs better in a given situation. We will use the following symbols: $N_\textrm{wlay}$ is the number of $w$-layers for $w$-stacking, $N_\textrm{pix}$ is the number of pixels in the image along each side, $N_\textrm{vis}$ is the number of visibilities, $N_\textrm{kern}$ is the size of the anti-aliasing kernel (see \S\ref{sec:gridding}), $N_\textrm{wkern}$ is the size of the $w$-kernel for $w$-projection, $w_{\max}$ is the maximum $w$ value and $\alpha_\textrm{FOV}$ is the imaging field of view.

@@

\section{The WSClean imager implementation} \label{sec:implementation}
With the purpose of testing new algorithms for the imaging of MWA data, we have written a new imager around the $w$-stacking algorithm. The imager is not specialized for the MWA, and has been successfully used for imaging data from other telescopes.

The new imager, called ``WSClean'', is written in the C++ language. It reads visibilities from CASA measurement sets and writes output images to Flexible Image Transport System (FITS) files. Several steps are multi-threaded using the threading module of the C++11 standard library. These are: reading and writing; gridding from and to different $w$-layers; performing the FFTs; and H\"ogbom Clean iterations (peak-finding and image subtraction, \citealt{hogbom-clean}). For the latter, intrinsics are used as well. Because of these optimizations, minor Cleaning iterations with an image size of $3072\times3072$ are performed at a rate of hundreds per second. This is fast enough to make the Clark Clean optimization \citep{clark-clean}, which consists of considering only a subset of pixels with a reduced PSF, less relevant. For very large images this optimization might still be useful, but we have not implemented it in WSClean. Because the PSF varies for imaging with non-zero $w$-values, subtracting a constant PSF in image space leads to inaccuracies. After a number of minor iterations, it is therefore beneficial to invert the model back to the visibilities via prediction, and subtract the model directly from the visibilities \citep{wprojection-cornwell}. This is similar to the Cotton-Schwab cleaning method \citep{cotton-schwab-clean}. WSClean allows cleaning individual polarizations, or can jointly deconvolve the polarizations. In the latter case, peaks are searched in $pp^2+2pq \overline{(pq)} + qq^2$ space, where $\overline{pq}=qp$ is the complex conjugate of $pq$, and the PSF is subtracted from the polarizations with different factors.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=13.5cm]{img/fullsky-together}
\caption{A beam-corrected MWA image of a 112~s zenith observation that covers almost the full sky. The lower image shows a zoom-in on the southern side lobe. PKS J2358-6054 ($\sim100$ Jy) is visible in the centre of the southern side lobe and resolved. The noise level in the side lobe is about 200 mJy. PKS J2358-6054 has been cleaned, but some artefacts remain because no direction-dependent calibration has been performed.}
\label{fig:full-sky-example}
\end{center}
\end{figure*}

@@

To efficiently image the full sky, an observation is split into short snapshots, their phase centre is changed to zenith and subsequently a suitable area is imaged. Depending on desired resolution and dynamic range, this might require very large images. To make an image at the MWA resolution, the image needs to have approximately 10k pixels along each side. The $w$-values will be very small at zenith, because at zenith only the vertical offsets of the antennas will contribute to the $w$-value. For the 128-tile MWA, the maximum differential elevation between tiles is $8.5$m. The tiles are in fact on a slight slope, and by fitting a plane to the antennas and changing the phase centre towards the normal of that plane, the maximum $w$-value decreases to $5.5\textrm{m} / \lambda$. In the following sections, when discussing the MWA zenith direction we are referring to this optimal $w$-direction.

@@

An example of the difference in projection between non-recentred and recentred imaging is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13-min MWA observation imaged with WSClean. The processing steps performed for this image were i) preprocessing of the observation with the Cotter preprocessing pipeline, which include time averaging and RFI flagging with the AOFlagger \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}; ii) calibrate on Hydra~A without direction dependence using a custom implementation of the RTS full-polarization calibration algorithm \citep{rts-mwa}; and iii) image each of the seven 112 s intervals separately. Cleaning an image that is in a different projection yields slightly different results, but qualitatively the two images are clearly of equal accuracy. The wall-clock time for imaging is 60 min and 41 min for respectively normal projection and the recentred image. Of the 41 min, 3 min is spent on phase rotating the visibilities.

Another method to shift an image is by using the periodicity of the FFT function. Since sources outside the field of view will be aliased back into the field, the image can be transformed to have the alias in the centre. This complicates gridding during imaging, because the anti-aliasing kernel needs to have a pass-band shape instead of a low-pass shape. Therefore, we chose to implement the former method.

\subsection{Beam correction}
\begin{figure*}
\begin{center}
\includegraphics[width=6cm]{img/pulsar-stokesi}\hspace{1cm}\includegraphics[width=6cm]{img/pulsar-stokesv}
\caption{Stokes I (left panel, total power) and Stokes V (right panel, circular polarization) images of PSR J0437-4715, demonstrating the full-polarization beam correction capability for MWA observations with WSClean. The pulsar (centre of image) displays 17\% circular polarization.}
\label{fig:pulsar-stokes-iv}
\end{center}
\end{figure*}
Because the MWA consists of fixed, beam-formed dipole antennas, the voltage beam of a MWA tile is described by a complex, non-diagonal Jones matrix. Alternatively, Mueller matrices can be used as well, but in general these result in more computations \citep{revisiting-me-i}. When assuming all individual antennas of the tiles are working properly, all tiles have the same beam. The beam can then be corrected in image space for all tiles at once with
\begin{equation}\label{eq:beam-correction}
 I(l, m) = J(l,m)^{-1} I'(l, m) \left(J(l,m)^*\right)^{-1},
\end{equation}
where each term is a $2\times2$ complex matrices, with $J$ the voltage beam and $^*$ denoting the conjugate transpose. Because $J$ is complex and non-diagonal, all 4 combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such. However, this is somewhat confusing because they are not necessarily orthogonal.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters. The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry. To make proper MWA beam correction possible, WSClean can output both real $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images. This requires four runs of the algorithm. Once these images are created, a separate program is used to perform the correction of Eq.~\eqref{eq:beam-correction}. The method is demonstrated in Fig.~\ref{fig:pulsar-stokes-iv}, which displays a detection of pulsar J0437-4715 in Stokes V. The image was made with an initial pipeline for the radio-sky monitor project that uses the MWA to search for transients and exoplanets. The pipeline includes WSClean to make the Stokes-parameter images.

@@

\subsection{Decreasing inversion resolution} \label{sec:decreasing-inversion-resolution}
Typically, for cleaning it is desired that images have a pixel size at least 5 times smaller than the synthesized-beam width. This improves the accuracy of the clean algorithm, because the positions of image maxima will be closer to the actual source positions. This factor of $\sim$5 is normally taken into account in the overall imaging resolution, i.e., the image size is increased during inversion. For the $w$-projection method, the cost of inversion with an increased resolution is small, because it does not increase the size of the $w$-kernels. It will affect the inverse FFT, but the relative cost of this step is negligible in the total cost. The $w$-stacking method is however significant affected by the image resolution, because it performs many inverse FFTs.

The spatial frequencies in the output image are band limited by the synthesized beam. Therefore, it is not required to perform the inversion at higher resolution. As long as the $uv$-plane is sampled with at least the Nyquist frequency, a high resolution image can be perfectly reconstructed from an inversion at lower resolution. After an up-scaled image has been cleaned, a model is created at the same (high) resolution. Because the model consists of delta functions, the spatial frequencies in the model are not band limited. Therefore, downscaling the model will remove information from the model. However, only the low spatial frequencies of this image will be used in the prediction. As long as downscaling does not modify the low-frequency components, the output of the prediction step will not change. This requires low-pass filtering the model before decimation.
